Principles for Open Data Curation: A Case Study with the New York City 311 Service Request Data

### 2024-09-26 manuscript/manuscript.tex
Review of sections 1 and 2
Responding to comments from JY inserted into manuscript text.

\documentclass[linenumber]{jdsart}

@@

\begin{document}

@@

\begin{abstract}
In the early 21st century, the open data movement began to transform 
societies and governments by promoting transparency,
innovation, and public engagement. New York City (NYC) has been at
the forefront of this movement since the enactment of the Open 
Data Law in 2012, which led to the creation of the NYC Open Data
portal. This portal now hosts 2,700 datasets from 80 city agencies,
serving as a crucial resource for research across various domains, 
including health, urban development, and transportation. The 
success of these initiatives highlights the importance of data 
curation in ensuring the utility and reliability of open datasets.
This paper examines the challenges of open data curation through a
case study of the NYC 311 Service Request (SR), addressing issues 
of data validity, consistency, and curation efficiency. Based on 
insights from this case study, we propose a set of data curation 
principles tailored for government-released open data. These principles 
aim to enhance data management practices and ensure 
the ongoing utility of open data. The paper concludes with 
actionable recommendations for enhancing data curation and outlines
general principles for the effective release of open data.

@@

In the early 21st century, the open data movement began 
to take shape, driven by the fundamental belief that 
freely accessible data can transform both societies and 
governments. This movement champions the principles
of transparency, innovation, and public engagement. 
A landmark in this journey was the launch of the United States'
\href{https://www.data.gov}{Data.gov} portal in 2009, a pioneering
platform in making government data widely accessible. Shortly after,
the European Union followed suit, unveiling its
\href{https://data.europa.eu/euodp}{Open Data Portal} in 2012, further
cementing the movement's global reach. Furthermore, the World Bank's Open
Data initiative, initiated in 2010, stands out as a comprehensive
repository for global development data, available at
\href{https://data.worldbank.org}{World Bank Open Data}. 
These initiatives represent significant strides in democratizing data, 
breaking down barriers that once kept valuable information 
on government performance in silos. Their collective impact 
extends beyond mere data sharing to fostering a culture of openness 
that benefits individuals, communities, governments, and economies worldwide 
\citep{barns2016mine, wang2016adoption}.

New York City (NYC) has emerged as a leader in the open data movement,
marked by the enactment of the Open Data Law in 2012
\citep{zuiderwijk2014open}. This landmark legislation led to the
creation of the \href{https://opendata.cityofnewyork.us}{NYC Open Data
  portal}, which today hosts an impressive array of 2,700 datasets
from 80 different city agencies. This resource has become invaluable
for researchers across various fields and has significantly enhanced
local government transparency. Popular datasets include information on
restaurant health inspection violations, car crashes, high school and
college enrollment statistics, jail inmate charges, and the location
of city-wide free Internet access points. These datasets have been
applied in civil life in various ways, such as mapping car crashes
involving pedestrians and visualizing high school and college
enrollment trends. Furthermore, they have enabled significant research
across multiple domains, including health \citep{cantor2018facets,
  shankar2021data}, urban development \citep{neves2020impacts}, and
transportation \citep{gerte2019understanding}, aiding in the
understanding and addressing of complex urban challenges.

@@

Research into data curation has explored these challenges in depth. 
Among the earliest discussions, \citet{witt2009constructing} focused 
on developing data curation profiles tailored to specific contexts, 
setting a precedent for targeted data management strategies. 
Addressing broader challenges in data sharing and management, 
\citet{borgman2012conundrum} highlighted the complexities of 
research data distribution, emphasizing the need for robust 
strategies. This is complemented by \citet{hart2016ten}, who outlined 
essential principles for effective data management, particularly 
emphasizing meticulous curation practices. The utility of curated 
open data is vividly illustrated in public health and global challenges, 
where \citet{cantor2018facets} demonstrated the utility of curated 
data in evaluating community health determinants, and 
\citet{shankar2021data} observed its critical role during the 
COVID-19 pandemic in managing collective responses.

@@

The paper is organized as follows. Section~\ref{sec:data} provides 
an overview of the history of the NYC 311 SR system and presents 
a summary of SR counts over the 10-year period from 2014 to 2023. 
Section~\ref{sec:issues} delves into specific data cleansing 
challenges affecting data quality and curation efficiency, 
including structural problems, adherence to the data dictionary, 
and the presence of missing, blank, or N/A entries. This section 
also investigates field compliance with legal or acceptable values, 
highlights logical inconsistencies, and examines concerning patterns 
in the data. We explore the balance between precision and accuracy 
and identify duplicate or redundant fields, along with observations 
on the Data Dictionary. Section~\ref{sec:recommendations} offers 
practical recommendations for mitigating or resolving these issues, 
while Section~\ref{sec:discussion} encapsulates key insights and 
discusses the broader implications of our findings.

\section{NYC 311 SR Data} 
\label{sec:data}
The NYC 311 service, a critical component of New York City's public
engagement and service response framework, serves as a centralized hub
for non-emergency inquiries and requests. Introduced in 2003, the NYC
311 system was designed to streamline the city's response to
non-emergency issues, ranging from noise complaints to street
maintenance requests. Initially a phone-based call center, the system
evolved into a comprehensive data management platform handling
millions of requests annually. Key milestones since its launch in 2003
include the addition of online and mobile app channels in 2009, a
\jy{should we mention the system upgrade just before the pandemic to
  document this historical event?}
record high of 348,463 monthly service requests in August 2020 due to
the COVID-19 pandemic, the 2021 expansion to include the MTA's city
subway system, and a record 3.23 million service requests in
2023. Today, the NYC 311 data system manages over 3 million service
requests per year. This data is publicly accessible through the NYC Open Data
Portal, which provides tools for querying, grouping, aggregating,
geo-mapping, visualizing, and exporting results. The
\href{https://www.nyc.gov/content/oti/pages/}{NYC Office of Technology
  and Innovation (OTI)}, which provides technological support for the
open data system.

\begin{comment}
Despite its success, the 311 system faces several challenges:
data timeliness, accuracy, and consistency, difficulties in
correlating data over long periods, excluding personally
identifiable information (PII), integration with stand-alone systems
at selected NYC agencies, and managing API usagefor numerous
third-party users. These challenges are addressed by various agency
open data managers and 
\end{comment}

@@

Our investigation focused on a 10-year period (2014--2023), downloaded
on \jy{Give date because the data is updated daily.}
This version of the data is available as a supplement to facilitate
reproducibility of the analysis presented in the sequel.
\jy{How many total? file size? Somewhere we used only two years that
  need to be clarified.}
Figure~\ref{fig:10-yr} shows the annual counts of NYC 311 SR over the
10-year period from 2014 to 2023, revealing a clear upward trend in
public engagement with the system. The most significant increase
occurred in 2020, when the number of requests surged to 3.23 million
due to the COVID-19 pandemic, illustrating the essential role of the
311 system in supporting residents during times of crisis. Overall, SR
counts increased by nearly 49.5\% between 2014 and 2023, reflecting
growing public reliance on the system for addressing non-emergency
issues. The steady rise in SRs can be attributed to the increased
accessibility of the 311 system via online and mobile platforms, as
well as heightened public awareness of the service. While the spike in
2020 was exceptional, the broader trend indicates sustained growth in
service request volume over the decade, highlighting the system's
evolving role in managing both routine city operations and
extraordinary events.

\jy{The 'title' in each figure could be removed because we have
  captions. The fonts of the annotations in the figures need to be
  increased. The code generating them need to be tidied for future
  sharing.
}

\jy{The plot sizes seem to be fixed in the generation code. That's why
  the fonts are small when scaled. I'd generate them with approxiately
  the expected sizes in the manuscript. I'd also use aspect ratio
  0.618 (golden ratio).
}

Figure~\ref{fig:SRcountbyAgency} provides a breakdown of SRs by
agency, showing the cumulative percentage of SRs handled by each
agency over a given time period. This figure highlights that the
distribution of SRs is concentrated among a few key agencies. The six
largest agencies are New York Police Department (NYPD), 
Housing Preservation and Development (HPD),
New York City Department of Sanitation (DSNY), Department of
Transportation (DOT), Department of Environmental Protection (DEP),
and Department of Parks and Recreation (DPR).
They collectively handle over 90\% of the total SR volume. This is
indicative of the critical roles these agencies play in managing
public concerns, ranging from noise complaints and housing issues to
sanitation and transportation problems. The NYPD alone accounts for a
substantial portion of the total SRs, underscoring its central role in
addressing complaints related to public safety and order. The
remaining 10\% distributed across a large number of smaller
agencies. This concentration of requests points to the systemic
reliance on a few key departments for resolving most non-emergency
issues, further emphasizing the importance of optimizing data handling
and curation processes within these high-volume agencies to ensure
efficiency and responsiveness.

@@

\begin{comment}
\begin{table}[tbp]
  \centering
  \caption{Noise-related complaints\_type(s) by count with Agency}
  \label{tab:noisecomplaints}
  \begin{tabular}{@{}lS[table-format=7.0,round-mode=places,
    round-precision=0]S[table-format=2.2,round-mode=places,
    round-precision=2]l@{}} % 'l' for left-aligned, 'S' for siunitx number alignment
    \toprule
    \textbf{complaint\_type} & \textbf{Count} & \textbf{Percentage} & \textbf{Agency} \\ 
    \midrule
    Noise - Residential        & 675502 & 10.56 & NYPD  \\ 
    Noise - Street/Sidewalk    & 300507 &  4.70 & NYPD  \\ 
    Noise - Commercial         & 128892 &  2.02 & NYPD  \\ 
    Noise - Vehicle            & 115956 &  1.81 & NYPD  \\ 
    Noise                      & 100413 &  1.57 & DEP   \\ 
    Noise - Helicopter         &  85345 &  1.33 & EDC   \\ 
    Noise - Park               &  16620 &  0.26 & NYPD  \\ 
    Noise - House of Worship &   2757 &  0.04 & NYPD  \\ 
    \bottomrule
  \end{tabular}
\end{table}
\end{comment}

\begin{figure}[tbp]
 \centering
  \includegraphics[width=0.7\textwidth]{SR_by_Complaint_Type.pdf} 
  \caption{Top 20 complaint type(s) and Cumulative Percentage} 
  \label{fig:SR_complaints}
\end{figure}

Figure~\ref{fig:SR_complaints}
illustrates the top 20 complaint types, accounting for 70\% 
of all Service Requests (SRs), and their cumulative percentage. The 
distribution of complaints is skewed, with a small number of issues 
dominating the total SR volume. Noise-related complaints, including 
residential, commercial, and street noise, make up 22\% of all 
complaints, making them the most frequent issue in the NYC 311 system. 
Other prominent categories include illegal parking, heat/hot water 
complaints, and blocked driveways. The cumulative percentage curve 
shows that after the top 20, the remaining complaint types are spread 
thinly across the remaining 30\% of SRs. This concentration suggests 
that improving responses to the most frequent complaints could have 
a significant impact on overall service efficiency and resident 
satisfaction. It also provides insights into the operational 
pressures faced by the city agencies responsible for handling these 
high-volume complaints.

@@

Data cleansing refers to the process of identifying and rectifying
errors, inconsistencies, and inaccuracies within datasets to ensure
they are of high quality and reliable for analysis
\citep{maletic2005data, hosseinzadeh2023data}. The process
typically involves removing duplicate records, handling missing or
incomplete data, correcting mislabeled or inaccurate entries, and
standardizing data formats \citep[e.g.,][]{cody2017cody,
  van2018statistical}. In the context of open data, cleansing is
especially important as open datasets often come from diverse,
uncoordinated sources, leading to variations in data quality,
completeness, and consistency. Without proper cleansing, the utility
of open data can be severely limited, affecting its reliability for
research, policy-making, and innovation. The main purpose of cleansing
open data is to ensure that it is accurate, consistent, and usable
across multiple platforms and by various stakeholders. This improves
the trustworthiness of the data and enables better decision-making,
more accurate analysis, and the integration of data into machine
learning models or other systems.

% What is data cleansing?  Wikipedia 
% \href{https://en.wikipedia.org/wiki/Data_cleansing}{Data Cleansing} 
% offers this definition.

% \begin{quote}\textit{Data cleansing or data cleaning is the process of detecting and 
% correcting (or removing) corrupt or inaccurate records from a record set, 
% table, or database and refers to identifying incomplete, incorrect, 
% inaccurate or irrelevant parts of the data and then replacing, 
% modifying, or deleting the dirty or coarse data.}
% \end{quote}

Many quality criteria are employed to ensure high-quality data 
processing. One of the primary efforts is data validation, which spans 
several critical checks. For instance, mandatory fields must not be 
left empty, ensuring that key information is always captured. 
Additionally, certain fields must conform to specific data types, 
such as numeric, character, or date formats, which are typically 
outlined in a Data Dictionary. Another essential aspect of validation 
is domain compliance, where fields must adhere to a predefined set of 
values, such as statuses, state names, zip codes, or gender. Structural 
errors also play a significant role in data quality, particularly when 
naming conventions or data entries are inconsistent. Common issues 
include fields that do not appear in the Data Dictionary or the 
inconsistent use of blanks, spaces, NA, N/A, or ``<NA>'' to indicate 
missing data. Furthermore, redundant or irrelevant fields can clutter 
datasets, reducing efficiency. Logical inconsistencies are another 
important consideration, such as related fields that violate expected 
relationships, like a ``due date'' that precedes the ``created date.'' 
Lastly, the balance between accuracy and precision is crucial, as both 
must be carefully managed to ensure reliable and meaningful data.

Here we identify the presence of issues in the NYC 311 SR open data;
not necessarily to solve them. A solution effort should
be undertaken only after an investigation as to the why and how 
such issues came about, and a discussion as to whether or not it 
is an \textit{error}.

@@

\begin{itemize}
	\item There are 47 columns of data for each row, exportable as a CSV file.

@@

	\item There are three borough fields; two of which appear to be duplicates.

	\item Two zip code fields, but not duplicates

@@

	\item Two Police Precinct fields; not duplicates

@@

The 311 SR 
\href{https://data.cityofnewyork.us/api/views/erm2-nwe9/files/b372b884-f86a-453b-ba16-1fe06ce9d212?download=true&filename=311_ServiceRequest_2010-Present_DataDictionary_Updated_2023.xlsx}{Data Dictionary} 
identifies 41 data columns (fields) and provides related information 
for each. However, upon downloading the data, six additional fields 
appear that are not included in the Data Dictionary. These fields are 
visible in the portal's Column Manager widget, labeled with the prefix 
``@computed\_region.'' The additional fields include \texttt{zip\_codes}, 
\texttt{community\_districts}, \texttt{borough\_boundaries}, 
\texttt{city\_council\_districts}, \texttt{police\_precincts}, and 
\texttt{police\_precinct}. While one can infer the meaning and 
derivation of these computed fields to some extent, uncertainty remains. 
The critical question is whether these fields can be reliably used for 
analytical purposes reliable documentation. In particular, which one
of the two percinct fields should one use? What is the difference
between the computed zip codes and the incident zipcode among the 41
documented variables?

@@

\paragraph{Zip Codes}
\label{sec:zipcodesissues}
All zip codes (two fields: \texttt{zip\_codes} and \texttt{incident\_zip}) 
should be valid as defined by the USPS database, which contains 
37,946 valid zip codes. However, the computed field \texttt{zip\_codes} 
proved problematic, with over 58\% (3.6 million) of its entries being 
invalid. Although the more accurate \texttt{incident\_zip} field has 
only 0.07\% invalid entries, this still results in 4,163 errors, which 
is not insignificant. The breakdown of invalid entries in the 
\texttt{zip\_codes} field, sorted by Agency, shows that the
\jy{Which agency? Missing a table? We could prepare a supplement where
  we can put additional tables/figures without limit.}
distribution by percentage mirrors the overall breakdown of SRs by 
Agency, potentially indicating a systemic problem.

@@

%\begin{figure}[tbp]
%	 \centering
%	 \includegraphics[width = \textwidth]{zero_duration_SR.pdf}
%	 \caption{SRs with Zero Durations by Agency}
%	 \label{fig:zero-duration}
%\end{figure}

@@

@@

Table~\ref{tab:x1nearmatches} provides examples of near-matches for 
\texttt{cross\_street\_1} and \texttt{intersection\_street\_1}. These 
entries show slight variations in the street name or number, such as 
``WEST 168 ST'' and ``WEST 167 ST,'' both processed by DOT, where only 
one character differs. Although these discrepancies may seem small, 
they can have implications for data accuracy, particularly when 
geocoding or matching locations for analysis purposes.

@@

Engagement with data scientists and statisticians is key to improving 
data curation strategies. These experts can help design and refine 
curation processes, such as creating algorithms for automatic data 
cleaning, identifying inconsistent data entries, and ensuring the 
dataset’s structure is maintained over time. Their expertise is also 
critical in developing protocols for validating incoming data, correcting 
errors in real-time, and ensuring consistency across multiple data sources. 
Feedback from statisticians and data scientists, such as the issues reported 
in this paper, needs a clear and accessible channel to reach the 311 SR data 
team and other relevant city agencies. Establishing such feedback loops 
ensures that recommendations for improving data curation and validation 
are not only heard but implemented. This feedback mechanism would also allow 
ongoing dialogue between data users and city officials, leading to more dynamic 
and responsive data management practices.

@@

\section*{Supplementary Material}
The NYC 311 SR data and code used to produce the results reportred in
this paper are available [FixMe].
\jy{need to think: we need to keep a copy of the downloaded data;
  if the data portal changes and our code may not work.}

>>>>

\documentclass[linenumber]{jdsart}
 \usepackage{setspace}

@@

\doublespacing
\begin{document}

@@

\begin{abstract}
In the early 21st century, the open data movement began to transform 
societies and governments by promoting transparency,
innovation, and public engagement. The City of New York (NYC) has been at
the forefront of this movement since the enactment of the Open 
Data Law in 2012, leading to the creation of the NYC Open Data
portal. This portal now hosts 2,700 datasets from 80 city agencies,
serving as a crucial resource for research across various domains, 
including health, urban development, and transportation. The 
success of these initiatives highlights the importance of data 
curation in ensuring the utility and reliability of open datasets.

This paper examines the challenges of open data curation through a
case study of the NYC 311 Service Request (SR) dataset, addressing issues 
of data validity, consistency, and curation efficiency. Based on 
insights from this case study, we propose a set of data curation 
principles tailored for government-released open data. These principles 
aim to enhance data management practices and ensure 
the ongoing utility of open data. The paper concludes with 
actionable recommendations for enhancing data curation and outlines
general principles for the effective release of open data.

@@

In the early 21st century, the open data movement began 
to take shape, driven by the fundamental belief that 
freely accessible data can transform both societies and 
governments. This movement champions the principles
of transparency, innovation, and public engagement. 
A landmark in this journey was the launch of the United States'
\href{https://www.data.gov}{Data.gov} portal in 2009, a pioneering
platform in making government data widely accessible. Shortly after,
the European Union followed suit, unveiling its
\href{https://data.europa.eu/euodp}{Open Data Portal} in 2012, further
cementing the movement's global reach. Furthermore, the World Bank's Open
Data initiative, initiated in 2010, stands out as a comprehensive
repository for global development data, available at
\href{https://data.worldbank.org}{World Bank Open Data}. 
These initiatives represent significant strides in democratizing data, 
breaking down barriers that once kept valuable information 
regarding government performance in silos. Their collective impact 
extends beyond mere data sharing to fostering a culture of openness 
that benefits individuals, communities, governments, and economies worldwide 
\citep{barns2016mine, wang2016adoption}.

The City of New York (NYC) has emerged as a leader in the open data movement,
marked by the enactment of the Open Data Law in 2012
\citep{zuiderwijk2014open}. This landmark legislation led to the
creation of the \href{https://opendata.cityofnewyork.us}{NYC Open Data
  portal}, which today hosts an impressive array of 2,700 datasets
from 80 different city agencies. This resource has become invaluable
for researchers across various fields and has significantly enhanced
local government transparency. Popular datasets include information on
restaurant health inspection violations, car crashes, high school and
college enrollment statistics, jail inmate charges, and the location
of city-wide free Internet access points. These datasets have been
applied in civil life in various ways, such as mapping car crashes
involving pedestrians and visualizing high school and college
enrollment trends. Furthermore, they have enabled significant research
across multiple domains, including health \citep{cantor2018facets,
  shankar2021data}, urban development \citep{neves2020impacts}, and
transportation \citep{gerte2019understanding}, aiding in the
understanding and addressing of complex urban challenges.

@@

Research into data curation has explored these challenges in\mbox{-}depth. 
Among the earliest discussions, \citet{witt2009constructing} focused 
on developing data curation profiles tailored to specific contexts, 
setting a precedent for targeted data management strategies. 
Addressing broader challenges in data sharing and management, 
\citet{borgman2012conundrum} highlighted the complexities of 
research data distribution, emphasizing the need for robust 
strategies. This is complemented by \citet{hart2016ten}, who outlined 
essential principles for effective data management, particularly 
emphasizing meticulous curation practices. The utility of curated 
open data is vividly illustrated in public health and global challenges, 
where \citet{cantor2018facets} demonstrated the utility of curated 
data in evaluating community health determinants, and 
\citet{shankar2021data} observed its critical role during the 
COVID-19 pandemic in managing collective responses.

@@

The paper is organized as follows. Section~\ref{sec:data} provides 
an overview of the history of the NYC 311 SR system and presents 
a summary of SR counts over the 10-year period from 2014 to 2023. 
Section~\ref{sec:issues} delves into specific data cleansing 
challenges affecting data quality and curation efficiency, 
including structural problems, adherence to the data dictionary, 
and the presence of missing, blank, or N/A entries. This section
utilizes 311 SR data over a a 2-year period from 2022-2023.  It 
 investigates field compliance with reference or acceptable values, 
highlights logical inconsistencies, and examines concerning patterns 
in the data. We explore the balance between precision and accuracy 
and identify duplicate or redundant fields, along with observations 
on the Data Dictionary. Section~\ref{sec:recommendations} offers 
practical recommendations for mitigating or resolving these issues, 
while Section~\ref{sec:discussion} encapsulates key insights and 
discusses the broader implications of our findings.

\section{NYC 311 SR Data} 
\label{sec:data}
The NYC 311 service, a critical component of New York City's public
engagement and service response framework, serves as a centralized hub
for non-emergency inquiries and requests. Introduced in 2003, the NYC
311 system was designed to streamline the city's response to
non-emergency issues, ranging from noise complaints to street
maintenance requests. Initially a phone-based call center, the system
evolved into a comprehensive data management platform handling
millions of requests annually. Key milestones since its launch in 2003
include the addition of online and mobile app channels in 2009, a
record high of 348,463 monthly service requests in August 2020 due to
the COVID-19 pandemic, the 2021 expansion to include the MTA's city
subway system, and a record 3.23 million service requests in
2023.

Today, the NYC 311 data system manages over 3 million service
requests (SRs) per year. This data is publicly accessible through the NYC Open Data
Portal, which provides tools for querying, grouping, aggregating,
geo-mapping, visualizing, and exporting results. The
\href{https://www.nyc.gov/content/oti/pages/}{NYC Office of Technology
  and Innovation (OTI)}, provides technological support for the
open data system and data portal. Input data is sourced from 17 different
NYC Agencies.

\begin{comment}
Despite its success, the 311 system faces several challenges:
data timeliness, accuracy, and consistency, difficulties in
correlating data over long periods, excluding personally
identifiable information (PII), integration with stand-alone systems
at selected NYC agencies, and managing API usage for numerous
third-party users. These challenges are addressed by various agency
open data managers and 
\end{comment}

@@

Our investigation utilizes two 311 SR datasets covering 10-years and 2-years.
The 10-year period (2014--2023), downloaded on 15 September 2024, 
contains only three columns (created\_date, closed\_date, and 
agency). This dataset is used to analyze long-term 
trends and observe certain dating anomalies. The 10-year dataset is approximately 
1.4 GB in size and contains over 27 million rows, each row representing one SR.
 Instructions on downloading this file from the NYC Open Data Portal 
 are available in the supplemental material.

Additionally, this investigation uses a 2-year dataset covering the period
2022--2023, also downloaded on 15 September 2024. This data set 
contains the full 41 columns of 311 SR data available from the portal, 
and is used to conduct detailed analysis in areas such as data consistency, 
data validity, redundant fields, etc.. This 2-year dataset 
is quite large at 3.7GB. It contains over 6 million rows of SRs. Instructions 
on downloading this file from the NYC Open Data Portal are 
available in the supplemental material.

Figure~\ref{fig:10-yr} shows  annual counts of 311 SRs over the
10-year period from 2014 to 2023, revealing a clear upward trend in
public engagement. As indicated, the most significant increase
in activity occurred in 2020, when the number of requests surged to 3.23 million
during the COVID-19 pandemic, illustrating the essential role of the
311 system in supporting residents during times of crisis. Overall, SR
counts increased by 49.5\% between 2014 and 2023, reflecting
growing public reliance on the system. (During this 10-year time 
period, the population of NYC grew just 0.6\%.) The steady 
rise in SRs can be attributed to the increased accessibility of 
the 311 system via online and mobile platforms, as well as 
heightened public awareness  and utilization of the service. 
While the spike in 2020 was exceptional, the broader trend 
indicates sustained growth in service request volume over 
the decade, highlighting the system's evolving role in managing 
both routine city operations and extraordinary events.

\jy{The plot sizes seem to be fixed in the generation code. That's why
  the fonts are small when scaled. I'd generate them with approximately
  the expected sizes in the manuscript. I'd also use aspect ratio
  0.618 (golden ratio).
}

Figure~\ref{fig:SRcountbyAgency} provides a breakdown of SRs by
agency, showing the cumulative percentage of SRs handled by each
agency over a the 2-year time frame. Note that the
distribution of SRs is concentrated among a few key agencies. The six
largest agencies are New York Police Department (NYPD), 
Housing Preservation and Development (HPD),
New York City Department of Sanitation (DSNY), Department of
Transportation (DOT), Department of Environmental Protection (DEP),
and Department of Parks and Recreation (DPR).
Collectively they handle over 90\% of the total SR volume. This is
indicative of the critical roles these agencies play in managing
public concerns, ranging from noise complaints and housing issues to
sanitation and transportation problems. The NYPD alone accounts for a
substantial portion of the total SRs, underscoring its central role in
addressing complaints related to public safety. The
remaining 10\% is distributed across a number of other
agencies. This concentration of SRs points to a
reliance on a few key departments to resolve most non-emergency
issues (e.g. non 911 calls). Accordingly, this underscores 
the necessity for optimal data handling and curation processes 
within these high-volume agencies in order to ensure 
efficiency and responsiveness.

@@

\begin{figure}[tbp]
 \centering
  \includegraphics[width=0.7\textwidth]{SR_by_Complaint_Type.pdf} 
  \caption{Top 20 complaint types and Cumulative Percentage} 
  \label{fig:SR_complaints}
\end{figure}

Figure~\ref{fig:SR_complaints}
illustrates the top 20 complaint types, accounting for 70\% 
of all SRs and their cumulative percentage. Note that the 
distribution of complaints is skewed, with a small number of issues 
dominating total SR volume. Noise-related complaints, including 
residential, commercial, and street noise, make up 22\% of all 
complaints, making them the most frequent issue in the NYC 311 system. 
Other prominent categories include illegal parking, heat/hot water 
complaints, and blocked driveways. The cumulative percentage curve 
shows that after the top 20, the remaining complaint types are spread 
thinly across the remaining 30\% of SRs. This concentration suggests 
that improving responses to the most frequent complaints could have 
a significant impact on overall service efficiency and resident 
satisfaction. It also provides insights into the operational 
pressures faced by the city agencies responsible for handling these 
high-volume complaints.

@@

Data cleansing refers to the process of identifying and rectifying
errors, inconsistencies, and inaccuracies within datasets to ensure
they are of high quality and reliable for analysis
\citep{maletic2005data, hosseinzadeh2023data}. The process
typically involves removing duplicate records, handling missing or
incomplete data, correcting mislabeled or inaccurate entries, and
standardizing data formats \citep[e.g.,][]{cody2017cody,
  van2018statistical}. In the context of open data, cleansing is
especially important as open datasets often come from diverse,
uncoordinated sources, leading to variations in data quality,
completeness, and consistency. Without thorough data cleansing, the utility
of open data can be quite limited, affecting its reliability for
research, policy-making, and innovation. The main purpose of cleansing
open data is to ensure that it is accurate, consistent, and usable
across multiple platforms and by various stakeholders. This improves
the trustworthiness of the data and enables better decision-making,
more accurate analysis, and the integration of data into machine
learning models or other systems.

Many quality criteria are employed to ensure high-quality data 
processing. One of the primary efforts is data validation, which spans 
several critical checks. For instance, mandatory fields must not be 
left empty, ensuring that key information is always captured. 
Additionally, certain fields must conform to specific data types, 
such as numeric, character, or date formats, which are typically 
outlined in a Data Dictionary. Another essential aspect of validation 
is domain compliance, where fields must adhere to a predefined set of 
values, such as statuses, state names, zip codes, or gender. Structural 
errors also play a significant role in data quality, particularly when 
naming conventions or data entries are inconsistent. Common issues 
include fields that do not appear in the Data Dictionary or the 
inconsistent use of blanks, spaces, NA, N/A, or ``<NA>'' to indicate 
missing data. Furthermore, redundant or irrelevant fields can clutter 
datasets, reducing efficiency. Logical inconsistencies are another 
important consideration, such as related fields that violate expected 
relationships, such as a ``due date'' that precedes the ``created date.'' 
Lastly, the balance between accuracy and precision is crucial, as both 
must be carefully managed to ensure reliable and meaningful data.

Here we identify the presence of issues in the NYC 311 SR open data;
not attempting to solve them. A solution effort should
be undertaken only after an investigation as to the why and how 
such issues came about, and a discussion as to whether or not it 
is an actual \textit{error} or holds some other status.

@@

\begin{itemize}
	\item There are 41 columns of data for each row, exportable as a CSV file.
	Each row represents one Service Request.

@@

	\item There are two borough fields; one of which appear to be a duplicate.

@@

@@

The 311 SR 
\href{https://data.cityofnewyork.us/api/views/erm2-nwe9/files/b372b884-f86a-453b-ba16-1fe06ce9d212?download=true&filename=311_ServiceRequest_2010-Present_DataDictionary_Updated_2023.xlsx}{Data Dictionary} 
identifies 41 data columns (fields) and provides related information 
for each. However, there are six additional fields that can be
optionally downloaded. These fields are 
visible in the portal's ``Column manager'' widget, labeled with the prefix 
``@computed\_region.'' The additional fields include \texttt{zip\_codes}, 
\texttt{community\_districts}, \texttt{borough\_boundaries}, 
\texttt{city\_council\_districts}, \texttt{police\_precincts}, and 
\texttt{police\_precinct}. While one can likely infer the meaning and 
derivation of these computed fields, uncertainty remains as to
their derivation and usage. These six fields are not included in
the Data Dictionary. Our analysis has shown that these six 
fields have numerous data validity issues. Liaison with the 
NYC Open Data Teams leads us to treat these six fields 
as experimental and not for suitable for normal use. Accordingly, 
we will not address these fields in our analysis.

@@

\paragraph{Zip Codes}
\label{sec:zipcodesissues}
All zip codes (two fields: \texttt{zip\_codes} and \texttt{incident\_zip}) 
should be valid as defined by the USPS database, which contains 
37,946 valid zip codes. The \texttt{incident\_zip} field has 
only 0.07\% invalid entries, this still results in 4,163 errors; 
not insignificant. If you sort the invalid entries by Agency, 
the distribution by percentage mirrors the \texttt{overall} 
breakdown of SRs by Agency, potentially 
indicating a systemic problem.

@@

@@

A unique way to visualize this anomaly is to examine the busiest day 
during this 2-year period (Friday, 2023-09-29). Then aggregate 
the created\_date(s) by minute (with seconds equal to zero), providing
a minute-by-minute look at SR creation on that one day. Note the spike 
at exactly noon (12:00), well beyond the 3\textsigma line.

\begin{figure}[tbp]
    \centering
    \begin{minipage}{0.5\textwidth}
        \centering
        \includegraphics[width=\textwidth]{2-year-trend_SR_created_by_minute_of_busiest_day.pdf}
        \caption{SRs Created Minute-by-Minute on Busiest Day}
        \label{fig:busiestcreated}
    \end{minipage}%
    \begin{minipage}{0.5\textwidth}
        \centering
        \includegraphics[width=\textwidth]{2-year-trend-SR_closed_by_minute_of_busiest_day.pdf}
        \caption{SRs Closed Minute-by-Minute on Busiest Day}
        \label{fig:busiestclosed}
    \end{minipage}
\end{figure}

@@

Table~\ref{tab:x1near-matches} provides examples of near-matches for 
\texttt{cross\_street\_1} and \texttt{intersection\_street\_1}. These 
entries show slight variations in the street name or number, such as 
``WEST 168 ST'' and ``WEST 167 ST,'' both processed by DOT, where only 
one character differs. Although these discrepancies may seem small, 
they can have implications for data accuracy, particularly when 
geocoding or matching locations for analysis purposes.

@@

Engagement with data scientists and statisticians is key to improving 
data curation strategies. These experts can help design and refine 
curation processes, such as creating algorithms for automatic data 
cleaning, identifying inconsistent data entries, and ensuring the 
dataset(s) structure is maintained over time. Their expertise is also 
critical in developing protocols for validating incoming data, correcting 
errors in real-time, and ensuring consistency across multiple data sources. 
Feedback from statisticians and data scientists, such as the issues reported 
in this paper, needs a clear and accessible channel to reach the 311 SR data 
team and other relevant city agencies. Establishing such feedback loops 
ensures that recommendations for improving data curation and validation 
are not only heard but implemented. This feedback mechanism would also allow 
ongoing dialogue between data users and city officials, leading to more dynamic 
and responsive data management practices.

@@

\section*{Supplementary Material}
The NYC 311 SR data and code used to produce the results reported in
this paper are available [FixMe].
\jy{need to think: we need to keep a copy of the downloaded data;
  if the data portal changes and our code may not work.}
